---
title: "Lecture 2 MATH 390.4 Queens College"
author: "Professor Adam Kapelner"
date: "January 31, 2018"
---

Let's talk about some simple operations on vectors:

```{r}
x = seq(1, 100) 
length(x) #how long is this vector
head(x) #the first elements
tail(x) #the last elements
head(x, 11) #the first 11 elements
tail(x, 13) #the last 13 elements
x[1] #R has one-based indexing while C, C++, Python, Java, have zero-based indexing
x[6]
x[length(x)] #unfortunately no "-1" last index convenience...
x[c(2, 77)]
x[2 : 77]
x[seq(1, 100, by = 2)] #odd numbers
x[-1] #everything but 1
x[-(1 : 10)] #everything but 1 - 10
x[-seq(2, 100, by = 2)] #another way to see odd numbers
x = c(rep(1, 10), rep(2, 10), rep(3, 10))
unique(x) #only unique elements (in order in which they appear)
```

and comparisons on vectors

```{r}
x = 1 : 5
y = 1 : 5
x == y
all.equal(x, y)
```

Most operations in R are vector-operations:

```{r}
x = 1 : 5
x + 2
x * 2
x^2

```


A note about logical vectors:

```{r}
x = c(TRUE, TRUE)
y = c(TRUE, FALSE)
x = c(FALSE, FALSE)
!x
x & y
x | y
x && y #vector AND - confusing - recommend not to use this
x || y #vector OR - confusing - recommend not to use this
as.numeric(x)
sum(x) #does the numeric conversion for you
any(x) #convenient function
all(x)
#sometimes a useful function
xor(TRUE, FALSE)
xor(TRUE, TRUE)
xor(FALSE, FALSE)
```


Sampling is very important:

```{r}
x = 1 : 100
sample(x, 3)
sample(x, 101)
sample(x, 3, replace = TRUE)
sample(x, 101, replace = TRUE)
sample(x) #default is length of the vector i.e. just shuffles
```

There are many "illegal" computation values in R: NaN, Inf, -Inf,

```{r}
1 / 0 #unlike C, Java -- no error... handles this natively
1 / 0 == Inf #this works
0 / 0 #unlike C, Java -- no error... handles this natively
x = 0 / 0
x == NaN #beware!!
is.nan(x)
-1 / 0
1 + 1 / 0
1 + Inf
1 / Inf
1 / (-Inf)
1 / NaN
log(0)
log(-1)
sqrt(-1)
```

There are a couple more "illegalish" values: NA, NULL, empty vectors

```{r}
NA #this is the value of missing
x = NA
x == NA #beware!!
is.na(x)
x = NULL #special reserved primitive for data that is "undefined"
x == NULL #strange... and beware
is.null(x)
#let's look at that strange thing
x = 1 : 3
x[1]
x[0] #a vector with zero elements -- a "null vector" of sorts
length(x[0])
c() #remarkably... not what we expect...
x[NA]
x[NaN]
x[Inf]
```

Now let's look at data types again. These types are the data types we learned in class.

```{r}
x = c(1, 2, 3, 4, 5)
class(x)
x = seq(1, 5)
class(x) #integer and numeric are approximately the same for purposes of this class
x = sample(rep(c(0, 1), 50))
x #binary variable?
x = c("a", "b", "c", "d")
x
class(x)
x = rep(x, 5)
x
class(x)
x = factor(x)
x
?factor
levels = c("low", "medium", "high")
x_char = sample(rep(levels, 10))
x = factor(x_char)
x #nominal categorical variable
x = factor(x_char, levels = levels, ordered = TRUE)
x #ordinal categorical variable
as.numeric(x) #coerce this variable to a number... result makes sense
```

Data which is of class factor can be used in mathematical models that we build just like numeric. It will, by default be reduced to binary variables.

Let's now do some probability. Let's realize a few iid random variable models. Let's say $X_1, ..., X_{10} \iid$ ...

* binomial
* geometric
* poisson
* standard uniform --- all numbers between 0 and 1 equally likely
* exponential
* normal with mean 5 and standard deviation 2

```{r}
num_rvs = 10 #good programming practice to declare shared data

x = rbinom(num_rvs, size = 5, prob = 0.1)
x #returns as a vector
?rbinom
x = rgeom(num_rvs, prob = 0.1)
x
?rgeom
x = rpois(num_rvs, lambda = 2)
x
?rpois
x = runif(num_rvs)
x
?runif
x = rexp(num_rvs, rate = 1)
x
?rexp
x = rnorm(num_rvs, mean = 5, sd = 2)
x
?rnorm
```

Sorting is also a key skill in R:

```{r}
x = rbinom(50, size = 20, prob = 0.2)
x
sort(x)
sort(x, decreasing = TRUE)
```

Now that we finished vectors, let's learn about the matrix object. Let's use the dimension of the learning problem in class.

```{r}
n = 100 #number of historical objects: the people
p = 3 #number of features about each

random_junk = round(runif(n * p), 2)
random_junk
X = matrix(random_junk, nrow = n, ncol = p)
X
X[1, 1]
X[1, ] #first row
class(X[1, ]) #note: not matrix anymore
X[1, , drop = FALSE]
class(X[1, , drop = FALSE]) #still matrix... this matters for matrix computations (coming soon)
X[, 1] #first column
class(X[, 1]) #note: not matrix anymore
X[, 1, drop = FALSE]
class(X[, 1, drop = FALSE])
```

A few handy functions for numerical matrices:

```{r}
rowSums(X)
colSums(X)
rowMeans(X)
colMeans(X)
```

