---
title: "Lecture 4 MATH 390.4 Queens College"
author: "Professor Adam Kapelner"
date: "February 14, 2018"
---

## The Perceptron

Recall our fabricated data:

```{r}
n = 100
Xy = data.frame(
  salary = round(rnorm(n, 50000, 20000)),
  has_past_unpaid_loan = rbinom(n, size = 1, prob = 0.2),
  past_crime_severity = sample(
    c("no crime", "infraction", "misdimeanor", "felony"),
    size = n,
    replace = TRUE,
    prob = c(.50, .40, .08, .02)
  ),
  y = factor(rbinom(n, size = 1, prob = 0.9), labels = c("No", "Yes"))
)
row.names(Xy) = c(
  "Sophia", "Emma", "Olivia", "Ava", "Mia", "Isabella", "Riley", 
  "Aria", "Zoe", "Charlotte", "Lily", "Layla", "Amelia", "Emily", 
  "Madelyn", "Aubrey", "Adalyn", "Madison", "Chloe", "Harper", 
  "Abigail", "Aaliyah", "Avery", "Evelyn", "Kaylee", "Ella", "Ellie", 
  "Scarlett", "Arianna", "Hailey", "Nora", "Addison", "Brooklyn", 
  "Hannah", "Mila", "Leah", "Elizabeth", "Sarah", "Eliana", "Mackenzie", 
  "Peyton", "Maria", "Grace", "Adeline", "Elena", "Anna", "Victoria", 
  "Camilla", "Lillian", "Natalie", "Jackson", "Aiden", "Lucas", 
  "Liam", "Noah", "Ethan", "Mason", "Caden", "Oliver", "Elijah", 
  "Grayson", "Jacob", "Michael", "Benjamin", "Carter", "James", 
  "Jayden", "Logan", "Alexander", "Caleb", "Ryan", "Luke", "Daniel", 
  "Jack", "William", "Owen", "Gabriel", "Matthew", "Connor", "Jayce", 
  "Isaac", "Sebastian", "Henry", "Muhammad", "Cameron", "Wyatt", 
  "Dylan", "Nathan", "Nicholas", "Julian", "Eli", "Levi", "Isaiah", 
  "Landon", "David", "Christian", "Andrew", "Brayden", "John", 
  "Lincoln"
)
```

Now let's do the perceptron for one dimension - just salary. You will do the other two for homework.

```{r}
MAX_ITER = 1000
w_vec = c(0, 0)
y_binary = as.numeric(Xy$y == "Yes") #convenience

for (iter in 1 : MAX_ITER){  
  for (i in 1 : n){
    x_i = c(1, Xy[i, 1]) #prepend the 1
    yhat_i = ifelse(sum(x_i * w_vec) > 0, 1, 0)
    y_i = y_binary[i]
    w_vec = w_vec + (y_i - yhat_i) * x_i
  }
}
w_vec
```

## Computer Science Sidebar

Before we get back to modeling, it is worth knowing a couple more data structures in R. These are not "data [science] types", these are "[computer science] data types". 

The first are "lists" which are "ordered hashmaps" or "ordered dictionaries" or "hash tables". if You don't know what this is, you should read about this online as it should have been covered in a intro to CS class.

```{r}
dict = list()
dict$a = "first"
dict$b = "second"
dict$c = "third"
dict
length(dict)
names(dict) #keys
dict_unlisted = unlist(dict) #values
dict_unlisted
class(dict_unlisted) #i.e. a vector
#three ways to access values by key / ordered location
dict$a
dict[["a"]]
dict[[1]] #value of first entered key
class(dict[[1]])
dict[[1 : 2]] #bombs
#now let's try to access a value for a non-existent key / ordered location
dict$q
dict[["q"]]
dict[[4]] #bombs
#convenient means to subset the list
dict[1]
class(dict[1])
dict[1 : 2]
dict[1 : 4] #this is the reason this type of access is not recommended
dict = list("first", "second", "third") #no key => value... what happens?
dict #default keys are the numbers 1, 2, ...
dict[[1]]
```

Lists conveniently allow all sorts of data types.

```{r}
varied_dict = list()
varied_dict$a = "first"
varied_dict$b = 2
varied_dict$c = 1 : 7
varied_dict$d = matrix(NA, nrow = 2, ncol = 2)
varied_dict[["the last entry"]] = function(x){x^2} #this key is not recommended
varied_dict
varied_dict$`the last entry` #note the tick marks (sometimes seen) needed due to spaces in key name
length(varied_dict)
names(varied_dict)
```

They have lots of uses in data science applications. We will likely see them in class and if not, you'll definitely see them in the real world. Note that data.frame objects are implemented as lists as well as many other common R objects.


The second is arrays i.e. are multidimensional vectors

```{r}
x = array(1 : 5, 5)
x
X = array(1 : 25, dim = c(5, 5))
X
X = array(1 : 125, dim = c(5, 5, 5))
X
X[1, , ]
X[, 1, ]
X[, , 1]
X = array(1 : 125, dim = c(5, 5, 5))
X
```

These can be associative arrays too and operate like a hash of vectors across arbitrary dimensions:

```{r}
X = array(1 : 125, 
          dim = c(5, 5, 5),
          dimnames = list(
            c("A", "B", "C", "D", "E"),
            c("I", "II", "III", "IV", "V"),
            c("blue", "red", "green", "yellow", "orange")
          ))
X
X["A", , ]
X[, "III", ]
X[, , "orange"]
X["C", , "orange"]
X["C", "IV", "orange"]
```


## Saving and Loading in R

Let's save one object:

```{r}
save(varied_dict, file = "varied_dict.RData") #RData is recommended extension for convenience
rm(varied_dict)
varied_dict
load("varied_dict.RData")
varied_dict
```

Or save the whole workspace:

```{r}
save.image("all_work.RData")
rm(list = ls())
ls() #empty vector i.e. nothing here!
load("all_work.RData")
ls()
```

```{r}
rm(list = ls()) #cleanup everything
```
