---
title: "Lecture 7 MATH 390.4 Queens College"
author: "Professor Adam Kapelner"
date: "February 21, 2018"
---

First we load up the breast cancer data set.

```{r}
Xy = na.omit(MASS::biopsy) #The "breast cancer" data with all observations with missing values dropped
X = Xy[, 2 : 10] #V1, V2, ..., V9
y_binary = as.numeric(Xy$class == "malignant")
```

## Nearest Neighbor algorithm

In one dimension, we are looking for the closest x. Here is a simple function that does it:

```{r}
nn_function = function(x_star){
  best_sqd_distance = Inf #good place to begin
  i_star = NA
  for (i in 1 : nrow(X)){
    dsqd = (X[i, 1] - x_star)^2
    if (dsqd < best_sqd_distance){
      best_sqd_distance = dsqd
      i_star = i
    }
  }
  y_binary[i_star]
}
nn_function(7.8)
nn_function(5.2)
```

Why is this silly for this dataset?

```{r}
str(X)
```

The features are not truly continuous. Would it make sense in higher dimensions? Your homework...

Has this been coded before? Definitely...

```{r}
pacman::p_load(class)
?knn
```

We can fit a knn model *and* predict in one shot via:

```{r}
y_hat = knn(X, c(4, 2, 1, 1, 2, 1, 2, 1, 1), y_binary, k = 1)
y_hat
```

Why is build model and predict in one shot natural in knn?

Now for an interesting exercise that will setup future classes:

```{r}
y_hat = knn(X, X, y_binary, k = 1)
y_hat
all.equal(y_hat, factor(y_binary))
```

No errors! Can this be a good model? No... "something" must be wrong! It is too good to be true.

Something is wrong. This is the first example of "overfitting". We will explore this in depth.

```{r}
rm(list = ls())
```

## Simple Linear Regression

To understand what the algorithm is doing - best linear fit by minimizing the squared errors, we can draw a picture. First let's make up some very simple training data $\mathbb{D}$.

```{r}
n = 20
x = runif(n)
beta_0 = 3
beta_1 = -2
y = beta_0 + beta_1 * x + rnorm(n, mean = 0, sd = 0.33)
```

And let's plot the data:


```{r}
pacman::p_load(ggplot2)
simple_df = data.frame(x = x, y = y)
simple_viz_obj = ggplot(simple_df, aes(x, y)) + 
  geom_point(size = 2)
simple_viz_obj
```

And its true $h^*$ line:

```{r}
true_hstar_line = geom_abline(intercept = beta_0, slope = beta_1, color = "green")
simple_viz_obj + true_hstar_line
```
Now let's calculate the simple least squares coefficients:

```{r}
r = cor(x, y)
s_x = sd(x)
s_y = sd(y)
ybar = mean(y)
xbar = mean(x)

b_1 = r * s_y / s_x
b_0 = ybar - b_1 * xbar
b_0
b_1
```

Note how $b_0$ and $b_1$ are not exactly the same as $\beta_0$ and $\beta_1$.

And we can plot it:


```{r}
simple_ls_regression_line = geom_abline(intercept = b_0, slope = b_1, color = "red")
simple_viz_obj + simple_ls_regression_line + true_hstar_line
```

Review of the modeling framework:

The difference between the green line and red line is the "estimation error". The difference between the green line and the points is a combination of error due to ignorance and error due to misspecification of $f$ as a straight line. In most real-world applications, estimation error is usually small relative to the other two. In the era of "big data", $n$ is usually big so estimation error is pretty small.

Recall that the noise (epsilons) are the difference between the data and the green line:

```{r}
simple_df$hstar = beta_0 + beta_1 * simple_df$x
simple_viz_obj = ggplot(simple_df, aes(x, y)) + 
  geom_point(size = 2)
epsilon_line_segments = geom_segment(aes(xend = x, yend = hstar), position = position_nudge(x = 0.002))
simple_viz_obj + epsilon_line_segments + true_hstar_line
```

And that the residuals (e's) are the difference between the data and the green line:

```{r}
simple_df$gs = b_0 + b_1 * simple_df$x
simple_viz_obj = ggplot(simple_df, aes(x, y)) + 
  geom_point(size = 2)
e_line_segments = geom_segment(aes(xend = x, yend = gs), color = "purple")
simple_viz_obj + simple_ls_regression_line + e_line_segments
```

Examining both at the same time:

```{r}
simple_viz_obj + simple_ls_regression_line + true_hstar_line + e_line_segments + epsilon_line_segments
```
