---
title: "Lecture 1 MATH 390.4 Queens College"
author: "Professor Adam Kapelner"
---

This file type "Rmd" is "R Markdown" which allows for text and R interwoven together. 

You can get started with this cheatsheet:
https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf

Acknowledgement: I would like to thank Professor Andreas Buja at Wharton as I cribbed lots of his notes.


What is R?

* An open-source, free standard for statistical and numerical programming
* interpreted, scripted and high level (like Python and not like e.g. C++)
* Has a tremendous developer community and industry community
* Very good for "one-shot" scripting - very common in data science
* New stats algorithms and methodologies appear here first!
* I joke that I'm an R developer for a living
* History: came from Bell Labs. C --> S --> R

Let's see some R! Since R is interpreted, you can run commands directly into its command line:

```{r}
5
```

This is not the most exciting command. The above is called a "chunk" and it tells the R Markdown compiler that there is R code inside those tick marks.

Here's some R code commented:

```{r}
5 #This will produce the number 5
```

Everything beyond the "#" is a comment as is ignored during interpretation.

Let's start by using R as a calculator

```{r}
5 + 5
5 - 5
5 / 5
5 * 5
5^5 #or 5 ** 5 (not common)
5 / 3
5 %% 3
5 * 5 %% 3
(5 * 5) %% 3 #note order of operations - use parentheses
```

You can have multiple statements per line:

```{r}
5 + 5; 5 - 5
```

This is considered bad style... only do this when you are making a point.

You can also extend expressions for many lines

```{r}
5 +
  5 +
  5
```
Avoid this unless the line gets so long it's difficult to read. Note the "+" at the end of the line so the interpreter knows to keep looking to the next line.

Let's take a look at how to represent numbers using scientific notation

```{R}
3,125 #commas are illegal
3.125e3 #scientific notation
3.125E3 
3.125E-3
options(digits = 3)
3.125E-3
options(digits = 7) #the default
```

There are lots of options. Look at the help file:

```{r}
?options
```

Let's talk about data types:

```{r}
TRUE #logical
class(TRUE)
1 #numeric
class(1)
"A"
class("A")
```

All of the above are actually 1-dim vectors. We will explore this later.

Every language has a means of invoking functions. Here are some included famous math functions. R uses parentheses to invoke functions. The quantities within the parentheses are known as "parameters" or "arguments" to the function. Each of these functions takes in one argument:

```{r}
sqrt(2)
log(3)
log10(1000)
sin(pi) #pi is a hard-wired numerical constant, the only one in the language (I think)
#note the numerical error
acos(1)
exp(1)
```

To create vectors, we use the "c" function which is short for "combine". This function takes any number of arguments

```{r}
c(1, 1, 2, 3, 5, 8)
```

We can also use shorthand "ladder" notation:

```{r}
1 : 3
1 : 100
-3 : 3
3 : -3
-(1 : 5)
```

The "seq" function (short for "sequence") is key for this too:

```{r}
seq(1, 9) #short for 1 : 9
seq(1, 9, by = 2)
seq(1, 9, length.out = 3)
seq(1, 9, length.out = 4)
```

So is the "rep" function (short for "replicate"):

```{r}
rep(1, 10)
rep(c(1, 2, 3), 10)
```


Every language has a means to assign a piece of data to variables:

```{r}
x = 1 #I use this...
x
x <- 2 #but this is actually preferred according to the the R spec
x
```
