---
title: "Lecture 11 MATH 390.4 Queens College"
author: "Professor Adam Kapelner"
date: "March 12, 2018"
---

Let's load up the Boston Housing data and ggplot.

```{r}
boston = MASS::Boston
pacman::p_load(ggplot2)
```

##Extrapolation

Let's take a look at a model that explains median home value using the average number of rooms. We plot $g$ in blue.

```{r}
mod = lm(medv ~ rm, boston)
b = summary(mod)$coef
ggplot(boston, aes(rm, medv)) + 
  geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "blue", lwd = 2)
```

Let's plot this a little bit differently:

```{r}
pacman::p_load(ggplot2)
ggplot(boston, aes(rm, medv)) + 
  geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "blue", lwd = 2) +
  xlim(0, 13) + ylim(-10, 90)
```

What changed in this plot? Nothing but the limits of the x and y axes. And we can see more clearly that "lines are forever"... just like diamonds... except that the line is not so precious with `rm` less than 4 and `rm` greater than 9. What are the problems with this?

What did we learn here? This model really shouldn't be used outside of the range of $\mathbb{D}$. In this range, predictions are called "interpolations" amd pitseode pf the range they are called "extrapolations". The former is solid and latter is not. Why is this?

Is the linear model of median home value using rooms *actually* linear? Not a chance! You cannot have rooms be negative or more than e.g. 50. And you can't have median values that are negative. Linear models are usually only approximately linear in a sensical range of the predictors.


## Multivariate linear regression

We want to run a multivariate linear regression $\mathcal{H}$ employing the least squares $\mathcal{A}$ manually using our derived linear algebra. Let us first pull out $\mathbb{D}$ as $y$ and $X$.

Let's ensure we augment the `X` to include the 1 vector in front. We need this for the intercept in the $w$ vector in our spec, $\mathcal{H}$.

```{r}
y = boston$medv
X = cbind(1, boston[, 1: 13])
```

Can we find $X^\top X$?

```{r}
# XtX = t(X) %*% X
```

The data frame is great, but unfortunately R does not allow us to use matrix algebra on it.

So let's create a matrix. Note: there are no factor variables with more than one level. `chas` is a binary variable and that's okay. If there were factors with more than level, the following will not work. We will explore this later.

```{r}
X = as.matrix(cbind(1, boston[, 1: 13]))
```

So $p = 12$ and $p + 1 = 14$.

Let's make each predictor name nice just for aesthetic value:

```{r}
colnames(X)
colnames(X)[1] = "(intercept)" #this is the standard way lm denotes it (which we will compare to later)
colnames(X)
```


Can we find $X^\top X$?

```{r}
XtX = t(X) %*% X
```

Is it full rank?

```{r}
XtXinv = solve(XtX)
```

It worked. This means $X$ is full rank i.e. there is no linear duplication of information over the `13 + 1` predictors. In case we're in doubt:

```{r}
pacman::p_load(Matrix)
rankMatrix(X)[[1]]
rankMatrix(t(X))[[1]]
rankMatrix(XtX)[[1]]
rankMatrix(XtXinv)[[1]]
```


Let's calculate the LS solution then:

```{r}
b = XtXinv %*% t(X) %*% y
b
```

Interpretation: if `crim` "increases" by 1, $\hat{y}$ increases by... etc etc. How would `crim` increase? Big philosophical topic which we are punting on (for now). If all predictors are 0, then $y$ would be predicted to be the intercept, 20.65. Strange concept... not usually important.

What would $g$ look like?

```{r}
g_predict_function = function(x_star){
   x_star %*% b
}
g_predict_function(X[7, ])
y[7] #good prediction!
```

Pretty simple...  and `x_star` could be a matrix of `n_star * (p + 1)` - where `n_star` is however many new observations you wish to predict.

We can compute all predictions:

```{r}
yhat = X %*% b
```

Can you tell this is projected onto a 13 dimensionsal space from a 506 dimensional space? Not really... but it is...

We can calculate the residuals:

```{r}
e = y - yhat
head(e)
```

What is RMSE?

```{r}
SSE = t(e) %*% e
MSE = 1 / (ncol(X)) * SSE
RMSE = sqrt(MSE)
SSE
MSE
RMSE
```

Interpret the RMSE...

We can calculate $R^2$ two ways:

```{r}
s_sq_y = var(y)
s_sq_e = var(e)
Rsq = (s_sq_y - s_sq_e) / s_sq_y
Rsq

n = length(e)
SST = (n - 1) * s_sq_y
Rsq = 1 - SSE / SST
Rsq
```

Let's look at distribution of $y$ and $e$ to get an idea about $R^2$ as we did before:


```{r}
pacman::p_load(ggplot2)

ggplot(data.frame(null_residuals = y - mean(y), residuals = e)) + 
  stat_density(aes(x = residuals), fill = "darkgreen", alpha = 0.3) + 
  stat_density(aes(x = null_residuals, fill = "red", alpha = 0.3)) +
  theme(legend.position = "none")
```

What does this tell you about $R^2$?

Now, of course, R has its own function to do all this. We've already seen them! To run a multivariate least squares linear model,

```{r}
mult_lin_mod = lm(medv ~ ., boston)
```

No need to (a) create a matrix from the data frame (b) append a 1's column (c) do the linear algebra. It's all done for you. What is this formula `medv ~ .`? Previously we've seen `medv ~ rm` to indicate "fit phenomenon `medv` using predictor `rm`". Here, it's "fit phenomenon `medv` using all available predictors in the data frame". This is a very powerful formula!

Let's extract the estimates $b$ as well as $R^2$ and RMSE:

```{r}
coef(mult_lin_mod)
summary(mult_lin_mod)$r.squared
summary(mult_lin_mod)$sigma
```

Does R offer a simple way to do $g$? Sure...

```{r}
x_star = boston[7, ]
y_hat_star = predict(mult_lin_mod, newdata = x_star)
y_hat_star
y[7]
```

If you care about the internals of what R is doing, it retraces our steps perfectly. It first creates the "model matrix" we called the "design matrix" and denoted it X:

```{r}
Xmm = model.matrix(medv ~ ., boston)
head(Xmm)
head(X) #same
```

Then it uses an internal function to compute the linear algebra:

```{r}
raw_mod = lm.fit(Xmm, y)
raw_mod$coefficients
```

We will soon see the internals of the `lm.fit` algorithm when we get to orthogonal projections in class (coming up next).

```{r}
rm(list = ls())
```


## LS using categorical predictors

Note that historically this is called "Analysis of Variance" or "ANOVA" for short. But there is no difference to the computer, it still crunches the same matrices.

Let's get the cars data again:

```{r}
cars = MASS::Cars93
str(cars)
```

Let's try to model `Type`, a factor with 6 levels.

```{r}
table(cars$Type)
```

What will $\hay{y}$ look like? Should be the $\bar{y}$'s for each level. What is $p$? 6. Let' see:

```{r}
anova_mod = lm(Price ~ Type, cars)
coef(anova_mod)
summary(anova_mod)$r.squared
```

What's going on here? This one categorical variable got blown up into 5 features! How to interpret?? First need reference category. Which one is missing? `Compact`. So what is prediction for the compact type? The intercept. What is prediction of Large type? Intercept + Large, etc. How does this regression look. Let's see the model matrix:

```{r}
Xmm = model.matrix(Price ~ Type, cars)
head(Xmm, 20)
```

The predictor `Type` got "dummified". (Remember we spoke about this in lecture 2)? There are now 5 dummy variables each representing one of the levels and the reference level is omitted because it is accounted for in the intercept. Let's make sure this is exactly what's going on.

```{r}
y = cars$Price
Xt = t(Xmm) 
XtX = Xt %*% Xmm
XtXinv = solve(XtX)
b = XtXinv %*% Xt %*% y
b
yhat = Xmm %*% b
e = y - yhat
Rsq = (var(y) - var(e)) / var(y)
Rsq
```

And of course the coefficients and $R^2$ are identical to the output from `lm`.

If we want to do a more "pure ANOVA", we can get rid of the intercept and see the $\bar{y}$'s immediately. This is handled in R's formula designation by adding a zero:

```{r}
anova_mod = lm(Price ~ 0 + Type, cars)
coef(anova_mod)
```

Is this correct?

```{r}
mean(cars$Price[cars$Type == "Compact"])
mean(cars$Price[cars$Type == "Large"])
mean(cars$Price[cars$Type == "Midsize"])
mean(cars$Price[cars$Type == "Small"])
mean(cars$Price[cars$Type == "Sporty"])
mean(cars$Price[cars$Type == "Van"])
```

What does $R^2$ look like?

```{r}
summary(anova_mod)$r.squared
```

Too high!! What happened? The $R^2$ calculation in `lm` is not accurate without the intercept. Keep this in mind. 

What does the design matrx (model matrix) look like?

```{r}
Xmm = model.matrix(Price ~ 0 + Type, cars)
head(Xmm, 20)
```

Very similar. 

Regressions without an intercept are not recommended. Here's why. What if we were doing two factors? I want a linear model with both Type and Airbags:

```{r}
table(cars$AirBags)
```

Airags is another nominal categorical variable, this time with three levels.

We invoke the model as follows.

```{r}
anova_mod = lm(Price ~ Type + AirBags, cars)
coef(anova_mod)
summary(anova_mod)$r.squared
```

What are interpretations now? What is the "reference level"? It's actually two levels in one: Type = compact and Airbags = Driver \& Passenger. 

A deeper question: can we read off Type = Midsize and AirBags = none? No... this is a modeling "enhancement" we will discuss in a few lectures from now.

If we model it without an intercept,


```{r}
anova_mod = lm(Price ~ 0 + Type + AirBags, cars)
coef(anova_mod)
```

we only get $\bar{y}$'s for the first factor predictor crossed with the reference category of the second. So above `TypeCompact` refers to the average of Type = Compact and Airbags = Driver \& Passenger.

Now let's create a linear model using one categorical predictor and one continuous predictor. The combination is called for historical reasons "Analysis of Covariance" or "ANCOVA" for short.

Let's use `Tyoe` and `Horsepower`:

```{r}
ancova_mod = lm(Price ~ Type + Horsepower, cars)
coef(ancova_mod)
summary(ancova_mod)$r.squared
```

Interpretation of estimated coefficients? Why did $R^2$ increase? (We will be explaining this in detail in the next unit).

What's going on the design / model matrix?

```{r}
head(model.matrix(Price ~ Type + Horsepower, cars))
```

Same as model matrix with just `Type`. Since `Horsepower` is continuous, it doesn't get dummified to more features.

What if we went back to the `Type` regression, left out the intercept, dummified and added the intercept back in?

```{r}
Xmm = model.matrix(Price ~ 0 + Type, cars)
Xmm = cbind(1, Xmm)
head(Xmm)
```

Are the columns linearly independent? No ... so when we try to get the hat matrix,

```{r}
Xmm %*% solve(t(Xmm) %*% Xmm) %*% t(Xmm)
```

You can't invert a non-invertible matrix!!

What does R do when using the linear model function:

```{r}
coef(lm(cars$Price ~ 0 + Xmm))
```

SOMEHOW: it doesn't complain since it handles the non-invertibility (we don't know why...) but we do see that it's busted. Look at the coefficients! One is missing! What is it doing?? It's just arbitrarily dropping one (just like recommended at the end of lecture 10).
